---
output: 
  bookdown::html_document2:
     toc: true
---
```{r, setup 008-diversity-inside-custom-host.R, include=FALSE}
knitr::opts_knit$set(root.dir = '/home/rakhimov/projects/amplicon_nmr')
```

````{r echo=FALSE}
# Alpha and beta diversity in NMR samples ####
````

# Alpha and beta diversity in NMR samples 

````{r echo=FALSE}
## Introduction ####
````


## Introduction
This script performs alpha diversity analysis and PCA on NMR samples

````{r echo=FALSE}
## 1. Load necessary libraries. ####
````


## Load necessary libraries.

````{r}
library(vegan)
library(tidyverse)
library(ggrepel)
````
````{r echo=FALSE}
## 2. Specifying parameters and directory/file names. #### 
````


## Specifying parameters and directory/file names. 
Name of the folder with QIIME2 output:

````{r}
authorname<-"pooled"
````

Directories with input files:

````{r}
rdafiles.directory<-"./output/rdafiles"
rtables.directory<-file.path("./output/rtables",authorname)
````

Truncation level that we chose in QIIME2:

````{r}
truncationlvl<-"234" 
````

Single reads or paired reads (decided in QIIME2):

````{r}
read.end.type<-"single"
rare.status<-"rare"
filter.status<-"nonfiltered"
````

The taxonomic rank that was used for agglomeration:

````{r}
agglom.rank<-"OTU"
````

Import rarefied data (rds).

````{r}
ps.q.df.preprocessed.date_time<-"20260211_17_14_21" # ASV NMR
````

Import abundance table as an rds file (NOT rarefied): 

````{r}
ps.q.agg.date_time<-"20260211_17_01_07" # ASV
ps.q.agg<-readRDS(file=file.path(
  rdafiles.directory,
  paste(ps.q.agg.date_time,"phyloseq-qiime",authorname,agglom.rank,
        read.end.type, truncationlvl,"table.rds",sep = "-")))
````

Import metadata.

````{r}
custom.md<-readRDS(file.path(rdafiles.directory,"custom.md.rds"))
````

Specify paths and image formats.

````{r}
image.formats<-c("png","tiff")
````

Labels for plots:

````{r}
metric.labs=c('sobs'= case_when(agglom.rank == "OTU" ~ "Richness (Observed species)",
                                agglom.rank == "Genus" ~ "Richness (Observed genera)") ,
              'shannon' = "Shannon",
              'invsimpson' = "Inverse Simpson")
plot.metrics<-c("sobs","shannon",
                "invsimpson") # metrics to plot
div.indices<-c("sobs","shannon",
               "invsimpson")
````

If you compare NMRs or mice (inside host):

````{r}
host<-"NMR"
# host.class<-c("NMR"="naked mole-rat",
#               "mice"="mouse")
# comparison<-"age"
# comparison<-"sex"
# comparison<-"strain"
gg.title.taxon <- ifelse(agglom.rank=="OTU","(ASV level)",
                       paste0("(",agglom.rank," level)"))

host.labels <- c("NMR" = "*Heterocephalus glaber*")
````

Load the rarefied dataframe:

````{r}
ps.q.df.preprocessed<-readRDS(
  file.path(rdafiles.directory,paste0(
    paste(
      ps.q.df.preprocessed.date_time,
      paste0("ps.q.df.",rare.status),filter.status,agglom.rank,
      paste(names(host.labels),collapse = '-'),sep = "-"),
    ".rds")))
````

Load metadata with age information:

````{r}
custom.md<-readRDS(file.path(rdafiles.directory,"custom.md.ages.rds"))%>%
  filter(sequencing_type == "Naked mole-rat 16S rRNA gene sequencing")
````

Theme for alpha diversity plots:

````{r}
alpha.plot.theme<-theme(#plot.margin=unit(c(1,1,1,2), 'cm'),
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  axis.text.x = ggtext::element_markdown(size=18),
  axis.text.y = element_text(size=18),
  axis.title = element_text(size = 20),
  strip.text.x = element_text(size=20),
  plot.title = element_text(size = 27),
  legend.text = element_text(size = 20),
  legend.title = element_text(size = 25),
  panel.grid.minor = element_blank(),
  panel.grid.major = element_blank())
````

Theme for PCA scatter plots:

````{r}
scatter.plot.theme<-theme(
  axis.text.x = element_text(angle=0,size=20,hjust=1),
  axis.text.y = element_text(size=20),
  axis.title = element_text(size = 20),
  plot.title = element_text(size = 27),
  legend.text = element_text(size = 20),
  legend.title = element_text(size = 25),
  legend.position = "right",
  plot.caption = ggtext::element_markdown(hjust = 0, size=20),
  plot.caption.position = "plot",
  panel.grid.minor = element_blank(),
  panel.grid.major = element_blank())

````
````{r echo=FALSE}
## 3. Alpha diversity analysis. ####
````


## Alpha diversity analysis. 

````{r echo=FALSE}
### 3.1 Compute alpha diversity metrics. ####
````


### Compute alpha diversity metrics.

````{r}
all.div<-ps.q.df.preprocessed%>%
  group_by(Sample)%>%
  reframe(sobs=specnumber(Abundance), # richness (num of species)
          shannon=diversity(Abundance,index = "shannon"),
          # simpson=diversity(Abundance, index="simpson"),
          invsimpson=diversity(Abundance, index="invsimpson"), # inverse simpson
          tot=sum(Abundance))%>%
  group_by(Sample)%>%
  pivot_longer(cols=c(sobs,shannon,invsimpson),
               names_to="metric")%>%
  distinct()

````
````{r echo=FALSE}
### 3.2 Alpha diversity tests. ####
````


### Alpha diversity tests. 
Function that runs Kruskal-Wallis test and pairwise Wilcoxon tests:

````{r}
test_alpha_diversity<-function(div.indices, 
                               all.div, 
                               comparison,
                               custom.levels,
                               custom.md){
  kt.results<-data.frame(matrix(nrow = 2,ncol=length(div.indices)))
  colnames(kt.results)<-div.indices
  rownames(kt.results)<-c("statistic", "pvalue")
  combinations<-combn(custom.levels,2) # all unique pairwise combinations
  w.results<-data.frame(matrix(nrow = ncol(combinations),
                               ncol=length(div.indices))) # ncol(combinations) pairwise comparisons
  colnames(w.results)<-div.indices
  w.results<-array(dim = c(length(table(combinations[1,])), # nrows
                           length(table(combinations[2,])), # ncols
                           length(div.indices)), # num of 2D arrays (stacking) 
                   dimnames = list(NULL, NULL, div.indices))
  for (div.metric in div.indices) {
    metric.ds<-all.div%>%
      left_join(custom.md)%>%
      filter(metric==div.metric)%>%
      distinct()
    # perform kruskal-wallis test
    if(comparison=="age"){
      kt<-kruskal.test(value~agegroup,data=metric.ds)
      
    }else if (comparison=="sex"){
      kt<-kruskal.test(value~sex,data=metric.ds)
      
    }else if(comparison=="strain"){
      kt<-kruskal.test(value~class,data=metric.ds)
      
    }
    
    kt.results["statistic",div.metric]<- kt$statistic
    kt.results["pvalue",div.metric]<- kt$p.value
    # low pvalue indicates statistical difference between some groups
    # But which groups are different from others?
    # Perform pairwise wilcoxon test
    
    # NB: You must not run pairwise wilcoxon test unless kruskal wallis results 
    # are significant
    if(kt$p.value<0.05){
      if(comparison=="age"){
        w.test<-pairwise.wilcox.test(metric.ds$value,
                                     metric.ds$agegroup,
                                     p.adjust.method = "BH",
                                     exact=FALSE)
      }else if (comparison=="sex"){
        w.test<-pairwise.wilcox.test(metric.ds$value,
                                     metric.ds$sex,
                                     p.adjust.method = "BH",
                                     exact=FALSE)
        
      }else if(comparison=="strain"){
        w.test<-pairwise.wilcox.test(metric.ds$value,
                                     metric.ds$class,
                                     p.adjust.method = "BH",
                                     exact=FALSE)
      }
      
      w.results[,,div.metric]<-w.test$p.value
      
    }else(
      w.results[,,div.metric]<-matrix(data = "n.s.",
                                      nrow = length(table(combinations[1,])), # nrows,
                                      ncol = length(table(combinations[2,])))
                                      # nrow = nrow(w.test$p.value),
                                      # ncol = ncol(w.test$p.value))
      
    )
    # dimnames(w.results)[[1]]<-dimnames(w.test$p.value)[[1]] # change rownames of w.results
    # dimnames(w.results)[[2]]<-dimnames(w.test$p.value)[[2]] # change colnames of w.results
    dimnames(w.results)[[1]]<-custom.levels[1] # change rownames of w.results
    dimnames(w.results)[[2]]<-custom.levels[2] # change colnames of w.results
  }
  return(list(kt.results = kt.results,
         w.results = w.results))
}

````
````{r echo=FALSE}
### 3.3 Test on ages: ####
````


### Test on ages:  

````{r}
custom.levels <- names(table(custom.md$agegroup))
alpha.test.age<-test_alpha_diversity(div.indices = div.indices, 
                     all.div = all.div, 
                     comparison = "age",
                     custom.levels = custom.levels,
                     custom.md = custom.md)
alpha.test.age$kt.results
alpha.test.age$w.results
# stopifnot(all(alpha.test.age$kt.results[2,]<0.05))
all(alpha.test.age$kt.results[2,]<0.05)
````
````{r echo=FALSE}
### 3.4 Test on sexes: ####
````


### Test on sexes:

````{r}
custom.levels <- names(table(custom.md$sex))
alpha.test.sex<-test_alpha_diversity(div.indices = div.indices, 
                     all.div = all.div, 
                     comparison = "sex",
                     custom.levels = custom.levels,
                     custom.md = custom.md)
alpha.test.sex$kt.results
alpha.test.sex$w.results
# stopifnot(all(alpha.test.sex$kt.results[2,]<0.05))
all(alpha.test.sex$kt.results[2,]<0.05)
````

For plots.

````{r}
max.values<-all.div%>%
  group_by(metric)%>%
  summarise(max_val=max(value))

````
````{r echo=FALSE}
### 3.5 Prepare data for plotting. ####
````


### Prepare data for plotting.

````{r}
custom.md<-custom.md%>%
  mutate(agegroup = factor(agegroup,levels = c("agegroup0_10", 
                                               "agegroup10_16")))%>%
  mutate(sex = case_when(grepl("Female",custom.md$sex,ignore.case = T) ~ "female",
                         grepl("Male",custom.md$sex,ignore.case = T) ~ "male"),
         sex = factor(sex,levels = c("female", "male")))


````
````{r echo=FALSE}
### 3.6 Plot diversity on sexes. ####
````


### Plot diversity on sexes.

````{r}
pretty.level.names.sex <-
  c("female" = "Females",
    "male" = "Males")
custom.levels.sex <- names(pretty.level.names.sex)
pretty.level.names.sex <-pretty.level.names.sex[which(names(pretty.level.names.sex)%in%custom.levels.sex)]
gg.labs.name.sex <-"Host sex"
gg.title.groups.sex <-"groups"

div.plot.sex<-all.div%>%
  filter(metric%in%plot.metrics)%>%
  left_join(custom.md)%>%
  ggplot(aes(x=factor(sex, levels=custom.levels.sex),
             y=value,
             fill=factor(sex)))+
  geom_boxplot(show.legend = FALSE,outliers = F)+
  facet_wrap(~factor(metric, # reorder facets
                     levels=plot.metrics),
             ncol=length(plot.metrics),
             scales="free_y", # free y axis range
             labeller = as_labeller(metric.labs))+ # rename facets
  theme_bw()+ 
  labs(fill = gg.labs.name.sex)+
  scale_fill_viridis_d(name = NULL,
                       option = "C", 
                       labels = unname(pretty.level.names.sex),
                       end = 0.9,
                       direction = (-1),
                       alpha = 0.5)+
  scale_color_manual(breaks = unname(pretty.level.names.sex),
                     labels=unname(pretty.level.names.sex))+
  scale_x_discrete(labels=pretty.level.names.sex,
                   limits=custom.levels.sex)+ # rename boxplot labels (x axis)
  # scale_fill_manual(values = custom.fill)+
  # ggtitle(paste("Alpha diversity of the gut microbiota of different",as.character(host.class[host]),gg.title.groups,
  #               gg.title.taxon))+
  alpha.plot.theme + # theme for alpha diversity plots
  theme(legend.position = "none",
        strip.text.x = element_text(size=15))

div.plot.sex.jitter<-div.plot.sex+
  geom_jitter(width = 0.1, # jitter spread
              shape = 1, # empty dots
              size=3,
              show.legend = FALSE)
````
````{r fig.width=10, fig.height=6}
print(div.plot.sex.jitter)

div.plot.sex.dots<-div.plot.sex+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.6) # add dots
````
````{r fig.width=10, fig.height=6}
print(div.plot.sex.dots)

# for(image.format in image.formats){
#   ggsave(paste0("./images/diversity/alpha/",
#                 paste(paste(format(Sys.time(),format="%Y%m%d"),
#                             format(Sys.time(),format = "%H_%M_%S"),sep = "_"),
#                       "alpha-jitter",
#                       paste(plot.metrics,collapse = "-"),
#                       host,"sex",agglom.rank,truncationlvl,
#                       sep = "-"),".",image.format),
#          plot=div.plot.sex.jitter,
#          width=10, height=6,units="in",
#          # width = 6000,height = 3000, units = "px",
#          dpi=300,device = image.format)
#   ggsave(paste0("./images/diversity/alpha/",
#                 paste(paste(format(Sys.time(),format="%Y%m%d"),
#                             format(Sys.time(),format = "%H_%M_%S"),sep = "_"),
#                       "alpha-dots",
#                       paste(plot.metrics,collapse = "-"),
#                       host,"sex",agglom.rank,truncationlvl,
#                       sep = "-"),".",image.format),
#          plot=div.plot.sex.dots,
#          width=10, height=6,units="in",
#          # width = 6000,height = 3000, units = "px",
#          dpi=300,device = image.format)
# }

````
````{r echo=FALSE}
### 3.7 Plot diversity on ages. ####
````


### Plot diversity on ages.

````{r}
pretty.level.names.age<-names(table(custom.md$old_agegroup))
names(pretty.level.names.age)<-names(table(custom.md$agegroup))
custom.levels<-names(pretty.level.names.age)
gg.labs.name.age<-"Age group"
gg.title.groups.age<-"age groups"

nmr.sample.levels<-custom.md%>%
  filter(class=="NMR")%>%
  # mutate(age=round(time_length(sampling_date-birthday,unit="years")))%>%
  dplyr::select(Sample,age,class)%>%
  arrange(age)%>%
  distinct()%>%
  mutate(NewSample=paste0(Sample," (",age," y)"))%>%
  mutate(Sample=factor(Sample,levels=Sample),
         NewSample=factor(NewSample,levels=NewSample))

nmr.age.div.plot<-all.div%>%
  filter(metric%in%plot.metrics)%>%
  left_join(nmr.sample.levels)%>%
  ggplot(aes(x=age,y=value,fill=age))+
  geom_point(size=4,stat = "identity",colour = "white",shape=21)+
  facet_wrap(~factor(metric, # reorder facets
                     levels=plot.metrics),
             nrow=length(plot.metrics),
             scales="free_y", # free y axis range
             labeller = as_labeller(metric.labs))+ # rename facets
  scale_x_continuous(breaks = round(seq(min(nmr.sample.levels$age), 
                                        max(nmr.sample.levels$age), by = 1),1)) +
  scale_fill_viridis_c(option="D")+
  labs(fill="Age (years)",
       x="Age (years)")+
  theme_bw()+
  # ggtitle(paste0("Alpha diversity of the naked mole-rat gut microbiota across age\n",gg.title.taxon))+
  alpha.plot.theme + # general theme
  theme(axis.title.x = element_text(size=20),
        legend.title = element_text(vjust = 2))

````
````{r fig.width=11, fig.height=8}
print(nmr.age.div.plot)
# for(image.format in image.formats){
#   ggsave(paste0("./images/diversity/alpha/",
#                 paste(paste(format(Sys.time(),format="%Y%m%d"),
#                             format(Sys.time(),format = "%H_%M_%S"),sep = "_"),
#                       "alpha-per-age",
#                       paste(plot.metrics,collapse = "-"),
#                       "NMR","age",agglom.rank,truncationlvl,
#                       sep = "-"),".",image.format),
#          plot=nmr.age.div.plot,
#          width=11, height=8,units="in",
#          # width = 4000,height = 3000,units = "px",
#          dpi=300,device = image.format)
# }


````
````{r echo=FALSE}
## 4. PCA ####
````


## PCA
Convert df into wide format.

````{r}
ps.q.df.wide.pca<-ps.q.df.preprocessed%>%
  dplyr::select(all_of(c(agglom.rank,"Sample","Abundance")))%>%
  pivot_wider(names_from = all_of(agglom.rank), 
              values_from = "Abundance",
              values_fill = 0)%>%
  as.data.frame()
````

Colnames are OTUs and rownames are sample IDs.

````{r}
rownames(ps.q.df.wide.pca)<-ps.q.df.wide.pca$Sample
ps.q.df.wide.pca<-ps.q.df.wide.pca[,-1] # remove sample names
ps.q.df.wide.pca<-as.matrix(ps.q.df.wide.pca) # convert to matrix
# The first crucial step is to transform the compositional data:
# Use a log-ratio transformation, typically the centered log-ratio (clr)
# or isometric log-ratio (ilr) transformation.
# The clr transformation is often preferred for interpretability, 
# while ilr is used for statistical properties.
# We will use robust CLR (RCLR) with natural log
ps.q.df.wide.pca.tfm<-decostand(ps.q.df.wide.pca,method="rclr",logbase = exp)

#### >if you want to exclude specific samples
if(host=="NMR"){
  ps.q.df.wide.pca<-ps.q.df.wide.pca[-which(rownames(ps.q.df.wide.pca)%in%c("M40")),]
}else if(host=="mice"){
  ps.q.df.wide.pca<-ps.q.df.wide.pca[-which(rownames(ps.q.df.wide.pca)%in%c("mf_1","MSM343")),]
}
# remove ASVs that are zero because their samples are removed
ps.q.df.wide.pca<-ps.q.df.wide.pca[,which(colSums(ps.q.df.wide.pca)!=0)]
ps.q.df.wide.pca.tfm<-decostand(ps.q.df.wide.pca,
                                method="rclr",
                                logbase = exp)
####<

# calculate principal components
pca.q<-prcomp(ps.q.df.wide.pca.tfm)
str(pca.q)
dim(pca.q$x)

# reverse the signs
pca.q$rotation<- -1*pca.q$rotation

# display principal components (loadings)
head(pca.q$rotation)

# reverse the signs of the scores
pca.q$x<- -1*pca.q$x

# display the first six scores
head(pca.q$x)

#calculate total variance explained by each principal component
pca.q$sdev^2 / sum(pca.q$sdev^2)

#calculate total variance explained by each principal component
var_explained = pca.q$sdev^2 / sum(pca.q$sdev^2)

#create scree plot
qplot(seq_along(1:nrow(ps.q.df.wide.pca.tfm)), var_explained) + 
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Scree Plot") +
  ylim(0, 1)

````
````{r echo=FALSE}
### 4.1 PCA Plot. ####
````


### PCA Plot.

````{r}
PC1<-pca.q$x[,1]
PC2<-pca.q$x[,2]
perc.var<-round(100*summary(pca.q)$importance[2,1:2],2)

# PC1.df<-as.data.frame(PC1)%>%rownames_to_column("Sample")
PC1.df<-as.data.frame(PC1)%>%mutate(Sample=rownames(ps.q.df.wide.pca))
# PC2.df<-as.data.frame(PC2)%>%rownames_to_column("Sample")
PC2.df<-as.data.frame(PC2)%>%mutate(Sample=rownames(ps.q.df.wide.pca))

pca.plot<-PC1.df%>%
  left_join(PC2.df)%>%
  # left_join(custom.md)
  left_join(custom.md)



pca.plot.sex<-pca.plot%>%
  ggplot(aes(x = PC1, y = PC2,
             color = sex,
             fill = sex,
             shape = sex))+
  scale_shape_manual(values = 21:22,
                     labels = unname(pretty.level.names.sex))+
  scale_color_viridis_d(labels = unname(pretty.level.names.sex),
                        option = "C",
                        end = 0.9,
                        direction = (-1))+
  scale_fill_viridis_d(name = NULL,
                       option = "C", 
                       labels = unname(pretty.level.names.sex),
                       end = 0.9,
                       direction = (-1))+
  guides(fill = "none",
         color = "none")+
  stat_ellipse(geom = "polygon",
               level = 0.8,
               alpha=0.2,
               show.legend = FALSE)+
  labs(shape=gg.labs.name.sex)+
  geom_point(size = 5,
             color = "black" 
  ) +
  labs(x=paste0("PC1 (", perc.var[1], "%)"),
       y=paste0("PC2 (", perc.var[2], "%)"))+
  theme_bw()+
  geom_hline(yintercept=0,linetype="dashed", color = "grey")+
  geom_vline(xintercept=0,linetype="dashed", color = "grey")+
  # ggtitle(paste("PCA between different",as.character(host.class[host]),gg.title.groups, "\n",
  #               gg.title.taxon))+
  scatter.plot.theme+  # theme for scatter plots (PCA, PCoA, nMDS)
  theme(legend.title = element_text(vjust = 2))

pca.plot.age<-pca.plot%>%
    ggplot(aes(x = PC1, 
               y = PC2, 
               color = age,
               shape= class,
               fill = age))+
    scale_shape_manual(values = 21)+
    scale_fill_viridis_c(option = "D")+
    scale_color_viridis_c(option = "D")+
    guides(color = "none",
           shape = "none")+
    labs(fill="Age (years)")+
  geom_point(size = 5,
             color = "black" 
  ) +
  labs(x=paste0("PC1 (", perc.var[1], "%)"),
       y=paste0("PC2 (", perc.var[2], "%)"))+
  theme_bw()+
  geom_hline(yintercept=0,linetype="dashed", color = "grey")+
  geom_vline(xintercept=0,linetype="dashed", color = "grey")+
  # ggtitle(paste("PCA between different",as.character(host.class[host]),gg.title.groups, "\n",
  #               gg.title.taxon))+
  scatter.plot.theme+  # theme for scatter plots (PCA, PCoA, nMDS)
  theme(legend.title = element_text(vjust = 2))

````
````{r fig.width=11, fig.height=8}
print(pca.plot.sex)
````
````{r fig.width=11, fig.height=8}
print(pca.plot.age)
````

Label the samples

````{r}
pca.sex.labeled<-pca.plot.sex +
  ggrepel::geom_text_repel(aes(label=Sample),
                           show.legend = FALSE,size=7) # add labels to samples

pca.age.labeled<-pca.plot.age+
  # ggrepel::geom_text_repel(aes(label=paste0(Sample," (",age,")")),
  ggrepel::geom_text_repel(aes(label=paste0(Sample," (",age,")")),
                           max.overlaps = Inf,
                           show.legend = FALSE,
                           size=7) +# add labels to samples
  theme(legend.title = element_text(vjust = 2))

````
````{r fig.width=11, fig.height=8}
print(pca.sex.labeled)
````
````{r fig.width=11, fig.height=8}
print(pca.age.labeled)

# for(image.format in image.formats){
#   ggsave(paste0("./images/diversity/pca/",
#                 paste(paste(format(Sys.time(),format="%Y%m%d"),
#                             format(Sys.time(),format = "%H_%M_%S"),sep = "_"),
#                       "pca",
#                       host,
#                       comparison,agglom.rank,truncationlvl,
#                       sep = "-"),".",image.format),
#          plot=pca.plot,
#          width=11, height=8,units="in",
#          # width = 4000,height = 2500,units = "px",
#          dpi=300,device = image.format)
#   ggsave(paste0("./images/diversity/pca/",
#                 paste(paste(format(Sys.time(),format="%Y%m%d"),
#                             format(Sys.time(),format = "%H_%M_%S"),sep = "_"),
#                       "pca-labeled",
#                       host,
#                       comparison,agglom.rank,truncationlvl,
#                       sep = "-"),".",image.format),
#          plot=pca.labeled,
#          width=11, height=8,units="in",
#          # width = 4000,height = 2500,units = "px",units = "px",
#          dpi=300,device = image.format)
# }

````
````{r echo=FALSE}
### 4.2 Find ASVs that contribute to PCs. ####
````


### Find ASVs that contribute to PCs.

````{r}
aload<-abs(pca.q$rotation)
head(sweep(aload,2,colSums(aload),"/"))
colSums(sweep(aload, 2, colSums(aload), "/"))
colSums(aload)
aload[1,1]/colSums(aload)[1]

pc.df<-as.data.frame(sweep(aload,2,colSums(aload),"/")[,1:2])
lapply(pc.df,max)
max.ind<-lapply(pc.df,which.max)
pc.df[max.ind$PC1,]
pc.df[max.ind$PC2,]

pc1.loadings<-pc.df%>%
  rownames_to_column(var="OTU")%>%
  arrange(-PC1)%>%
  head(n=3)%>%
  left_join(ps.q.df.preprocessed)%>%
  distinct()%>%
  right_join(ps.q.agg[,c("Sample","OTU","Species")])%>%
  filter(class=="NMR")%>%
  arrange(-PC1,-Abundance)

pc2.loadings<-pc.df%>%
  rownames_to_column(var="OTU")%>%
  arrange(-PC2)%>%
  head(n=3)%>%
  left_join(ps.q.df.preprocessed)%>%
  distinct()%>%
  right_join(ps.q.agg[,c("Sample","OTU","Species")])%>%
  filter(class=="NMR")%>%
  arrange(-PC2,-Abundance)

rbind(pc1.loadings,pc2.loadings)%>%
  dplyr::select(Sample,OTU,Species,Abundance)%>%
  group_by(Species,OTU)%>%
  mutate(ASV= paste(Species,cur_group_id()))%>%
  ungroup%>%
  mutate(OTU=ASV,
         OTU=gsub(" ","_",OTU))%>%
  head


sessionInfo()
rm(list = ls(all=TRUE))
gc()
````

